---
title: "Depths' impact on prediction bands"
output: rmarkdown::html_document
bibliography: our-bib.bib
---

```{r, include = FALSE, eval = T}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r echo = F, message = F, warning = F, cache = F, eval = T}
library(dplyr)
library(tidyr)
devtools::load_all()
```

```{r eval = F, cache = F}
library(EpiModel)
library(dplyr)
library(tidyr)
```

# Overview

This vignette focusing on understanding how depth for functional/filamental data can aid in creating the best prediction bands for epidemic models. In this example, we present some SIR data (see [Comparing bands and assessing containment](https://skgallagher.github.io/EpiCompare/articles/comparing-bands-and-assessing-containment.html) for examples that can quickly apply to higher dimensions.), that contains interesting group structure, and we explore how local vs global depth creates different prediction bands.

# Data generation 

Here's our visualization of simulations (actually from 2 different models), but we can imagine that we believe that the true epidemic comes from a mixture of these two models. That is

$$
\text{Epidemic} \sim .5 \cdot \text{Model}_1 + .5 \cdot \text{Model}_2
$$
where 
$$
\text{Model}_1 = SIR(\beta = .1, \gamma = .03) \\
\text{Model}_2 = SIR(\beta = .15, \gamma = .05)
$$
```{r}
## first group -----------
beta <- .1
gamma <- .03

out <- simulate_SIR_agents(n_sims = 50,
                           n_time_steps = 100,
                           beta = beta, gamma = gamma,
                           init_SIR = c(950, 50, 0))

df_group <- out %>% group_by(sim) %>% 
  agents_to_aggregate(states = c("tI", "tR"),
                      min_max_time = c(0,100)) %>%
  rename(S = "X0", I = "X1", R = "X2") 

## second group ----------

beta <- .15
gamma <- .05

out2 <- simulate_SIR_agents(n_sims = 50,
                            n_time_steps = 100,
                            beta = beta, gamma = gamma,
                            init_SIR = c(950, 50, 0))

df_group2 <- out2 %>% group_by(sim) %>% 
  agents_to_aggregate(states = c("tI", "tR"),
                      min_max_time = c(0,100)) %>%
  rename(S = "X0", I = "X1", R = "X2") 

df_all <- rbind(df_group %>% mutate(id = "first"),
                df_group2 %>% mutate(id = "second"))
df_all_vis <- df_all %>% 
  mutate(id2 = factor(paste0(id, sim), levels = paste0(rep(c("first", "second"), each = 100),
                                     rep(1:100, length = 200)))) %>%
  mutate(id2 = as.numeric(id2)) 
```



```{r warning = F}

theme_set(theme_sir())

all_curves <- ggplot(df_all_vis) +
  geom_path(aes(x = S, y = I, z = R, color = id, group = id2),
            alpha = .1) + 
  coord_tern() +
  labs(color = "Group")

all_curves
```


# Creation of prediction bands

Our current framework through `geom_prediction_band` uses a global distance-based depth based on work from Geenens \& Nieto-Reyes's distance depth (directly explorably with the function `distance_depth_function`) [@Geenens2017]. Below you'll see that, the visualized prediction band using this global depth doesn't actually capture the area of the simplex we might want - specifically it seems to like the interior between these two groups a little bit more than what you and I might want. This is largely associated with the global nature of depth.

```{r cache = T}
global_depth_pb <- all_curves + 
  geom_prediction_band(data = df_all_vis,
                       aes(x = S, y = I, z = R, 
                          sim_group = id2),
                       pb_type = "delta_ball", conf_level = .6) +
  coord_tern()

global_depth_pb
```


# A local approach

Agostinelli \& Romanazzi (2011) proposed an approach to create local depth, and we have extended their ideas to Geenens \& Nieto-Reyes into a local distance based depth (directly explorably with the function `local_distance_depth_function`) [@Agostinelli2011]. Agostinelli \& Romanazzi (2011) recommended a localized parameter $\tau$ around the 20-30% quantile of the distance distribution - which we explore here.

## Tau selection

```{r eval = T, cache = T}
dist_mat <- df_all %>%
  arrange(t) %>% # just to be safe
  select(-t) %>%
  group_by(id, sim) %>% group_split() %>%
  dist_matrix_innersq_direction(position = 2:4, verbose = TRUE)

quantile(dist_mat,probs = 0:10/10) %>% t %>%
  data.frame(check.names = F) %>%
  DT::datatable(options = list(dom = 't'))

tau <- quantile(dist_mat,probs = .3)["30%"]

dist_mat %>% c() %>% data.frame(dist = .) %>%
  ggplot(aes(x = dist)) + geom_histogram(bins = 30) +
  geom_vline(xintercept = tau)
```

# Building an understanding

To gain a bit of understanding, we the below code allows us to explore the depth of each curve relative to global depth and local depth.

```{r}
local_depth <- df_all %>%
  arrange(t) %>% # just to be safe
  select(-t) %>%
  group_by(id, sim) %>%
  filament_distance_depth(depth_func = 
    function(dist_mat) {
      local_distance_depth_function(dist_mat, tau = tau)
      }, data_columns = c("S","I","R"))

global_depth <- df_all %>%
  arrange(t) %>% # just to be safe
  select(-t) %>%
  group_by(id, sim) %>%
  filament_distance_depth(depth_func = 
    function(dist_mat) {
      distance_depth_function(dist_mat)
      }, data_columns = c("S","I","R"))

df_all_ld <- df_all_vis %>% group_by(id, sim) %>%
  nest() %>% 
  ungroup() %>% mutate(local_depth = local_depth,
                       global_depth) %>%
  unnest(col = c(data)) %>%
  mutate(local_depth2 = cut(local_depth, breaks = 5),
         global_depth2 = cut(global_depth, breaks = 5)) 
```

```{r}
df_all_ld %>% 
  ggplot(aes(x = S, y = I, z = R, color = global_depth, group = id2)) +
  geom_path() +
  coord_tern() +
  labs(title = "Global Depth", color = "global depth")
```

```{r}
df_all_ld %>% 
  ggplot(aes(x = S, y = I, z = R, color = local_depth^2, group = id2)) +
  geom_path() +
  coord_tern() +
  labs(title = "Local Depth", color = "(local depth)^2") 
```

Here's a different prediction band based on local depth. It's slightly different
because we're filtering first (we may drop a few more than we want).

```{r}

local_cutoff <- df_all_ld %>% pull(local_depth) %>% quantile(probs = .4)

local_depth_pb <- all_curves + 
  geom_prediction_band(data = df_all_ld %>% filter(local_depth > local_cutoff),
                       aes(x = S, y = I, z = R, 
                          sim_group = id2),
                       pb_type = "delta_ball", conf_level = 1) +
  
  coord_tern()

grid.arrange(global_depth_pb + labs(title = "global-based prediction band"),
             local_depth_pb + labs(title = "local-based prediction band"),
             nrow = 1)
```


# References